\glossex{Seja haji.}{seja haji}{Sun shine.PRS}{,,The sun shines.''}
\glossex{Seja vi haji.}{seja vi haji}{Sun IPVF shine.PRS}{,,The sun is shining.''}
\glossex{Ore Seja haji.}{ore seja haji}{now Sun shine.PRS}{,,The sun is shining.''}
\glossex{Seja va hajit.}{seja va haji-t}{sun PFV shine-PST}{,,The sun shone.''}
\glossex{Seja ze haji.}{seja ze haji}{sun FUT shine}{,,The sun will shine.''}
\glossex{Seja vi hajit.}{seja vi haji-t}{sun IPVF shine-PST}{,,The sun has been shining.''}
\glossex{Seja haji rede}{seja haji rede}{sun shine.PRS again}{,,The sun is shining again.''}
\glossex{Seja ze haji zetay}{seja ze haji zetay}{sun FUT shine tomorrow}{,,The sun will shine tomorrow.''}
\glossex{Seja haji siro.}{seja haji siro}{sun shine.PRS bright.ADV}{,,The sun shines brightly.''}
\glossex{Siro Seja haji}{siro seja haji}{bright.ADJ sun shine.PRS}{,,The bright sun shines.''}
\glossex{Seja sormi ore.}{seja sormi ore}{sun sunrise.VT.PRS now}{,,The sun is rising now.''}
\glossex{Sormi ore.}{sormi ore}{sunrise.VI.PRS now}{,,The sun is rising now.''}
\glossex{Recha rujalaros rajit.}{recha rujalar-os raji-t}{all people-PL shout-PST}{,,All the people shouted.''}
\glossex{Recha rajit.}{recha raji-t}{all shout-PST}{,,All the people shouted.''}
\glossex{Noreja rajit.}{noreja raji-t}{some.people shout-PST}{,,Some of the people shouted.''}
\glossex{Vyele rujalaros rajit ka razi.}{vyele rujalar-os raji-t ka razi}{many people-PL shout-PST two times}{,,Many of the people shouted twice.''}
\glossex{Osaker rajit ka razi.}{osaker raji-t ka razi}{many.people shout-PST two times}{,,Many of the people shouted twice.''}
\glossex{Gusto rujalaros rajiri vicho.}{gusto rujalar-os rajiri vicho}{happy people-PL shout.PRS often}{,,Happy people often shout.''}
\glossex{Muchi ka͞ufit ared.}{muchi ka͞ufi-t ared}{kitten jump-PST upward.ADV}{,,The kitten jumped up.''}
\glossex{Muchi ka͞ufit on tasek.}{muchi ka͞ufi-t on tasek}{kitten jump-PST on table}{,,The kitten jumped onto the table.''}
\glossex{Myi vir muchi uneńit.}{myi vir muchi un-eńi-t}{1SG.POSS little cat away-go-PST}{,,My little kitten walked away.''}
\glossex{Aḱame osupi.}{aḱame osupi}{rain fall.PRS}{,,It's raining.''}
\glossex{Aḱame osupit.}{aḱame osupi-t}{rain fall-PST}{,,The rain came down.''}
\glossex{Muchi labi in aḱame.}{muchi labi in aḱame}{cat play.PRS in rain}{,,The kitten is playing in the rain.''}
\glossex{Aḱame aśtot.}{aḱame aśto-t}{rain stop-PST}{,,The rain has stopped.''}
\glossex{Aḱame ze aśtopi naári.}{akame ze aśtopi naári}{rain FUT stop soon}{,,Soon the rain will stop.''}
\glossex{Avi bochada aḱame ze aśtopi naári.}{avi bochada aḱame ze aśtopi naári}{have.PRS hope rain FUT stop soon}{,,I hope the rain stops soon.''}
\glossex{Mosno faji lot volu chet.}{mosno faj-i lo-t volu chet}{wild animal-PL live-PST sometime here}{,,Once wild animals lived here.''}
\glossex{Egla rav́at gepo.}{egl-a rav́a-t gepo}{3SG-F look.around-PST slow.ADV}{,,Slowly she looked around.''}
\glossex{Ti uneńi do!}{ti uneńi do}{2SG go.away IMP}{,,Go away!''}
\glossex{Uneńi do!}{Uneńi do}{go.away IMP}{,,Go away!''}
\glossex{Eni heme!}{eni heme}{go CHR}{,,Let's go!''}
\glossex{Ti vage eni.}{ti vage eni}{2SG HORT go}{,,You should go.''}
\glossex{Vage eni.}{vage eni}{HORT go}{,,You should go.''}
\glossex{Ze esi gusto per eni.}{ze esi gusto per eni}{FUT be happy to.do go}{,,I will be happy to go.''}
\glossex{Egli ze iéni naári.}{egli ze i-éni naári}{3SG.M FUT VEN-go soon}{,,He will arrive soon.''}
\glossex{Chidoyi moroza unj́aret.}{chido-yi moroza un-j́are-t}{child-POSS ball out-roll-PST}{,,The baby's ball has rolled away.''}
\glossex{Ka bovos vi tesuchi.}{ka bov-os vi tesuchi}{two boy-PL IPVF cooperate.PRS}{,,The two boys are working together.''}
\glossex{Ka bovos vi fari tesiko.}{ka bov-os vi fari tesiko}{two boy-PL IPVF work.PRS together.ADV}{,,The two boys are working together.''}
\glossex{Neb́i͞a ze osupi mireisro.}{neb́i͞a ze osupi mireisro}{mist FUT fall probably.PART}{,,This mist will probably clear away.''}
\glossex{Fote͞o flos dolozi in reso.}{fote͞o flo-s dolozi in reso}{lovely.ADJ flower-PL grow.PRS in everywhere}{,,Lovely flowers are growing everywhere.''}
\glossex{Noni vage vibi gepe.}{noni vage vibi gep-e}{1PL HORT eat.PRS slow.ADV-COMP}{,,We should eat more slowly.''}
\glossex{Ti va iént tar chid.}{ti va i-én-t tar chid}{2SG PFV VEN-go-PST too fast.ADV}{,,You have come too soon.''}
\glossex{Ti vipini rizo͞e chiwi.}{ti vipini rizo-e chiwi}{2SG must.PRS.AUX neat-COMP write}{,,You must write more neatly.''}
\glossex{Siniko kebuér varsi rizo onvojibo.}{siniko kebuér varsi rizo onvojibo}{wonderful palace stand.PRS exact.ADV opposite.ADV}{,,Directly opposite stands a wonderful palace.''}
\glossex{Henryi pelir lukat.}{henr-yi pelir luka-t}{henry-POSS dog lose.way-PST}{,,Henry's dog is lost.''}
\glossex{Pelir yi Henri lukat chyi.}{pelir yi henri luka-t chyi}{dog POSS henry lose.way-PST REFL}{,,Henry's dog is lost.''}
\glossex{Myi muche esi ruko.}{myi muche esi ruko}{1SG.POSS cat be.PRS black}{,,My cat is black.''}
\glossex{Myi muche ruko.}{myi muche ruko}{1SG.POSS cat black}{,,My cat is black.''}
\glossex{Loesa yi vir hima esi girefaro.}{loesa yi vir hima esi girefar-o}{doll POSS small girl be.PRS damage-ADJ}{,,The little girl's doll is broken.''}
\glossex{Machagi fahojo inuro.}{machagi fahojo inuro}{sleep.PRS loud.ADV usually.ADV}{,,I usually sleep soundly.''}
\glossex{Chidos jikit mito Jek.}{chid-os jiki-t mito jek}{child-PL run-PST back.ADV jack}{,,The children ran after Jack.''}
\glossex{Epi huhu mibozor labi.}{epi huhu mibozor labi}{can after school play}{,,I can play after school.''}
\glossex{Noni ent o kavala per tamoki.}{noni en-t o kavala per tamoki}{1PL go-PST to village for visit.PRS}{,,We went to the village for a visit.''}
\glossex{Noni iént o aḱumar.}{noni i-én-t o aḱumar}{1PL VEN-go-PST to river}{,,We arrived at the river.''}
\glossex{Va ingek a ti}{va ingek a ti}{PFV wait.PST on 2SG}{,,I have been waiting for you.''}
\glossex{Naberos set ra͞o nafrita.}{naber-os set rao na-frita}{camper-PL sit.PST around camp-fire}{,,The campers sat around the fire.''}
\glossex{Vir hima e il muchi set ne ne͞a mi.}{vir hima e il muchi set nea mi}{small girl and 3SG.POSS kitten sit.PST near.physically 1SG}{,,A little girl with a kitten sat near me.''}
\glossex{Chido ingek a il vapal ner ostro.}{chido ingek a il vapal ner ostro}{child wait.PST on 3SG.POSS father near.physically door}{,,The child waited at the door for her father.''}
\glossex{Kavala yi vyekka hima mant il muchi asweo tay.}{kavala yi vyekka hima man-t il muchi asweo tay}{village POSS old.SUPL girl lose-PST 3SG.POSS kitten previous day}{,,Yesterday the oldest girl in the village lost her kitten.''}
\glossex{Kavala yi vyekka hima yi muchi lukat asweo tay.}{kavala yi vyekka hima yi muchi lukat asweo tay}{village POSS old.SUPL girl POSS kitten lose.way-PST previous day}{,,Yesterday the oldest girl in the village lost her kitten.''}
\glossex{In che kavala ti uńatal?}{in che kavala ti uńatal}{in DEM village 2SG be.born-PST}{,,Were you born in this village?''}
\glossex{Lipe tyi brat leehi?}{lipe tyi brat leehi}{good.ADV 2SG.POSS brother dance}{,,Can your brother dance well?''}
\glossex{Sinit rujaler?}{sini-t rujaler}{leave-PST man}{,,Did the man leave?''}
\glossex{A ti tyi sora iéni?}{a ti tyi sora iéni}{for 2SG 2SG.POSS sister come}{,,Is your sister coming for you?''}
\glossex{Zetay ti epi iéni?}{zetay ti epi iéni}{tomorrow 2SG can.AUX come}{,,Can you come tomorrow?''}
\glossex{}{}{}{,,Have the neighbors gone away for the winter?''}
\glossex{}{}{}{,,Does the robin sing in the rain?''}
\glossex{}{}{}{,,Are you going with us to the concert?''}
\glossex{}{}{}{,,Have you ever travelled in the jungle?''}
\glossex{}{}{}{,,We sailed down the river for several miles.''}
\glossex{}{}{}{,,Everybody knows about hunting.''}
\glossex{}{}{}{,,On a Sunny morning after the solstice we started for the mountains.''}
\glossex{}{}{}{,,Tom laughed at the monkey's tricks.''}
\glossex{}{}{}{,,An old man with a walking stick stood beside the fence.''}
\glossex{}{}{}{,,The squirrel's nest was hidden by drooping boughs.''}
\glossex{}{}{}{,,The little seeds waited patiently under the snow for the warm spring sun.''}
\glossex{}{}{}{,,Many little girls with wreaths of flowers on their heads danced around the bonfire.''}
\glossex{}{}{}{,,The cover of the basket fell to the floor.''}
\glossex{}{}{}{,,The first boy in the line stopped at the entrance.''}
\glossex{}{}{}{,,On the top of the hill in a little hut lived a wise old woman.''}
\glossex{}{}{}{,,During our residence in the country we often walked in the pastures.''}
\glossex{}{}{}{,,When will your guests from the city arrive?''}
\glossex{}{}{}{,,Near the mouth of the river, its course turns sharply towards the East.''}
\glossex{}{}{}{,,Between the two lofty mountains lay a fertile valley.''}
\glossex{}{}{}{,,Among the wheat grew tall red poppies.''}
\glossex{}{}{}{,,The strong roots of the oak trees were torn from the ground.''}
\glossex{}{}{}{,,The sun looked down through the branches upon the children at play.''}
\glossex{}{}{}{,,The west wind blew across my face like a friendly caress.''}
\glossex{}{}{}{,,The spool of thread rolled across the floor.''}
\glossex{}{}{}{,,A box of growing plants stood in the Window.''}
\glossex{}{}{}{,,I am very happy.''}
\glossex{}{}{}{,,These oranges are juicy.''}
\glossex{}{}{}{,,Sea water is salty.''}
\glossex{}{}{}{,,The streets are full of people.''}
\glossex{}{}{}{,,Sugar tastes sweet.''}
\glossex{}{}{}{,,The fire feels hot.''}
\glossex{}{}{}{,,The little girl seemed lonely.''}
\glossex{}{}{}{,,The little boy's father had once been a sailor.''}
\glossex{}{}{}{,,I have lost my blanket.''}
\glossex{}{}{}{,,A robin has built his nest in the apple tree.''}
\glossex{}{}{}{,,At noon we ate our lunch by the roadside.''}
\glossex{}{}{}{,,Mr. Jones made a knife for his little boy.''}
\glossex{}{}{}{,,Their voices sound very happy.''}
\glossex{}{}{}{,,Is today Monday?''}
\glossex{}{}{}{,,Have all the leaves fallen from the tree?''}
\glossex{}{}{}{,,Will you be ready on time?''}
\glossex{}{}{}{,,Will you send this message for me?''}
\glossex{}{}{}{,,Are you waiting for me?''}
\glossex{}{}{}{,,Is this the first kitten of the litter?''}
\glossex{}{}{}{,,Are these shoes too big for you?''}
\glossex{}{}{}{,,How wide is the River?''}
\glossex{}{}{}{,,Listen.''}
\glossex{}{}{}{,,Sit here by me.''}
\glossex{}{}{}{,,Keep this secret until tomorrow.''}
\glossex{}{}{}{,,Come with us.''}
\glossex{}{}{}{,,Bring your friends with you.''}
\glossex{}{}{}{,,Be careful.''}
\glossex{}{}{}{,,Have some tea.''}
\glossex{}{}{}{,,Pip and his dog were great friends.''}
\glossex{}{}{}{,,John and Elizabeth are brother and sister.''}
\glossex{}{}{}{,,You and I will go together.''}
\glossex{}{}{}{,,They opened all the doors and windows.''}
\glossex{}{}{}{,,He is small, but strong.''}
\glossex{}{}{}{,,Is this tree an oak or a maple?''}
\glossex{}{}{}{,,Does the sky look blue or gray?''}
\glossex{}{}{}{,,Come with your father or mother.''}
\glossex{}{}{}{,,I am tired, but very happy.''}
\glossex{}{}{}{,,He played a tune on his wonderful flute.''}
\glossex{}{}{}{,,Toward the end of August the days grow much shorter.''}
\glossex{}{}{}{,,A company of soldiers marched over the hill and across the meadow.''}
\glossex{}{}{}{,,The first part of the story is very interesting.''}
\glossex{}{}{}{,,The crow dropped some pebbles into the pitcher and raised the water to the brim.''}
\glossex{}{}{}{,,The baby clapped her hands and laughed in glee.''}
\glossex{}{}{}{,,Stop your game and be quiet.''}
\glossex{}{}{}{,,The sound of the drums grew louder and louder.''}
\glossex{}{}{}{,,Do you like summer or winter better?''}
\glossex{}{}{}{,,That boy will have a wonderful trip.''}
\glossex{}{}{}{,,They popped corn, and then sat around the fire and ate it.''}
\glossex{}{}{}{,,They won the first two games, but lost the last one.''}
\glossex{}{}{}{,,Take this note, carry it to your mother; and wait for an answer.''}
\glossex{}{}{}{,,I awoke early, dressed hastily, and went down to breakfast.''}
\glossex{}{}{}{,,Aha! I have caught you!''}
\glossex{}{}{}{,,This string is too short!''}
\glossex{}{}{}{,,Oh, dear! the wind has blown my hat away!''}
\glossex{}{}{}{,,Alas! that news is sad indeed!''}
\glossex{}{}{}{,,Whew! that cold wind freezes my nose!''}
\glossex{}{}{}{,,Are you warm enough now?''}
\glossex{}{}{}{,,They heard the warning too late.''}
\glossex{}{}{}{,,We are a brave people, and love our country.''}
\glossex{}{}{}{,,All the children came except Mary.''}
\glossex{}{}{}{,,Jack seized a handful of pebbles and threw them into the lake.''}
\glossex{}{}{}{,,This cottage stood on a low hill, at some distance from the village.''}
\glossex{}{}{}{,,On a fine summer evening, the two old people were sitting outside the door of their cottage.''}
\glossex{}{}{}{,,Our bird's name is Jacko.''}
\glossex{}{}{}{,,The river knows the way to the sea.''}
\glossex{}{}{}{,,The boat sails away, like a bird on the wing.''}
\glossex{}{}{}{,,They looked cautiously about, but saw nothing.''}
\glossex{}{}{}{,,The little house had three rooms, a sitting room, a bedroom, and a tiny kitchen.''}
\glossex{}{}{}{,,We visited my uncle's village, the largest village in the world.''}
\glossex{}{}{}{,,We learn something new each day.''}
\glossex{}{}{}{,,The market begins five minutes earlier this week.''}
\glossex{}{}{}{,,Did you find the distance too great?''}
\glossex{}{}{}{,,Hurry, children.''}
\glossex{}{}{}{,,Madam, I will obey your command.''}
\glossex{}{}{}{,,Here under this tree they gave their guests a splendid feast.''}
\glossex{}{}{}{,,In winter I get up at night, and dress by yellow candlelight.''}
\glossex{}{}{}{,,Tell the last part of that story again.''}
\glossex{}{}{}{,,''}
\glossex{}{}{}{,,Be quick or you will be too late.''}
\glossex{}{}{}{,,Will you go with us or wait here?''}
\glossex{}{}{}{,,She was always, shabby, often ragged, and on cold days very uncomfortable.''}
\glossex{}{}{}{,,Think first and then act.''}
\glossex{}{}{}{,,I stood, a little mite of a girl, upon a chair by the window, and watched the falling snowflakes.''}
\glossex{}{}{}{,,Show the guests these shells, my son, and tell them their strange history.''}
\glossex{}{}{}{,,Be satisfied with nothing but your best.''}
\glossex{}{}{}{,,We consider them our faithful friends.''}
\glossex{}{}{}{,,We will make this place our home.''}
\glossex{}{}{}{,,The squirrels make their nests warm and snug with soft moss and leaves.''}
\glossex{}{}{}{,,The little girl made the doll's dress herself.''}
\glossex{}{}{}{,,I hurt myself.''}
\glossex{}{}{}{,,She was talking to herself.''}
\glossex{}{}{}{,,He proved himself trustworthy.''}
\glossex{}{}{}{,,We could see ourselves in the water.''}
\glossex{}{}{}{,,Do it yourself.''}
\glossex{}{}{}{,,I feel ashamed of myself.''}
\glossex{}{}{}{,,Sit here by yourself.''}
\glossex{}{}{}{,,The dress of the little princess was embroidered with roses, the national flower of the Country.''}
\glossex{}{}{}{,,They wore red caps, the symbol of liberty.''}
\glossex{}{}{}{,,With him as our protector, we fear no danger.''}
\glossex{}{}{}{,,All her finery, lace, ribbons, and feathers, was packed away in a trunk.''}
\glossex{}{}{}{,,Light he thought her, like a feather.''}
\glossex{}{}{}{,,Every spring and fall our cousins pay us a long visit.''}
\glossex{}{}{}{,,In our climate the grass remains green all winter.''}
\glossex{}{}{}{,,The boy who brought the book has gone.''}
\glossex{}{}{}{,,These are the flowers that you ordered.''}
\glossex{}{}{}{,,I have lost the book that you gave me.''}
\glossex{}{}{}{,,The fisherman who owned the boat now demanded payment.''}
\glossex{}{}{}{,,Come when you are called.''}
\glossex{}{}{}{,,I shall stay at home if it rains.''}
\glossex{}{}{}{,,When he saw me, he stopped.''}
\glossex{}{}{}{,,Do not laugh at me because I seem so absent minded.''}
\glossex{}{}{}{,,I shall lend you the books that you need.''}
\glossex{}{}{}{,,Come early next Monday if you can.''}
\glossex{}{}{}{,,If you come early, wait in the hall.''}
\glossex{}{}{}{,,I had a younger brother whose name was Antonio.''}
\glossex{}{}{}{,,Gnomes are little men who live under the ground.''}
\glossex{}{}{}{,,He is loved by everybody, because he has a gentle disposition.''}
\glossex{}{}{}{,,Hold the horse while I run and get my cap.''}
\glossex{}{}{}{,,I have found the ring I lost.''}
\glossex{}{}{}{,,Play and I will sing.''}
\glossex{}{}{}{,,That is the funniest story I ever heard.''}
\glossex{}{}{}{,,She is taller than her brother.''}
\glossex{}{}{}{,,They are no wiser than we.''}
\glossex{}{}{}{,,Light travels faster than sound.''}
\glossex{}{}{}{,,We have more time than they.''}
\glossex{}{}{}{,,She has more friends than enemies.''}
\glossex{}{}{}{,,He was very poor, and with his wife and five children lived in a little low cabin of logs and stones.''}
\glossex{}{}{}{,,When the wind blew, the traveler wrapped his mantle more closely around him.''}
\glossex{}{}{}{,,I am sure that we can go.''}
\glossex{}{}{}{,,We went back to the place where we saw the roses.''}
\glossex{}{}{}{,,"This tree is fifty feet high," said the gardener.''}
\glossex{}{}{}{,,I think that this train leaves five minutes earlier today.''}
\glossex{}{}{}{,,My opinion is that the governor will grant him a pardon.''}
\glossex{}{}{}{,,Why he has left the city is a mystery.''}
\glossex{}{}{}{,,The house stands where three roads meet.''}
\glossex{}{}{}{,,He has far more money than brains.''}
\glossex{}{}{}{,,Evidently that gate is never opened, for the long grass and the great hemlocks grow close against it.''}
\glossex{}{}{}{,,I met a little cottage girl; she was eight years old, she said.''}