\newpage

\section{Preface}

\begin{spacing}{1.1}

I am very glad you are reading this book! My name is \emph{Koolder mal
Erlehirni} and this book is a new project we started over two years ago. At that
moment it is mostly just a dictionary, but it will be accompanied with second
book, describing the rules and grammar of Andro language.

Usage of this book does require some knowledge in linguistics. We will start
with description of conventions used in the dictionary.

In this book only the transcription (romanization) is used, so you don't have to
know anything about our alphabets or syllabaries.

\bigskip

Good luck!

\section[Abbreviations and format]{Abbreviations and formatting}

Every word in this dictionary is presented to you in a standarized form, for
example:

\dictword{isdar}[ˈis.dar]
\dictterm{n}{half}

The bold text is a text in a transcription. In this dictionary, the
transcription Ziri is used. The next one is a phonemic representation using the
symbols from the International Phonetic Alphabet, IPA. As you may notice, this
dictionary is not using the full phonetic/allophonic descriptions, mostly
because the Andro language is not that strict and you may hear a bit different
phonemes in different dialects.

The next thing written in \emph{italics} is the abbreviation of what part of the
speech is represented by the dictionary term:

\begin{table}[h]
\begin{tabular}{ll}
\emph{n}    & noun           \\
\emph{v}    & verb            \\
\emph{adj}  & adjective          \\
\emph{pro}  & pronoun \\
\emph{part} & particle
\end{tabular}
\end{table}

Depending on the dictionary term you will see additional markings:

\dictword{archit}[ˈar.ʈ͡ʂit]
\dictterm{n}{(\textsc{pl} architji \xm{ˈar.ʈ͡ʂit.ʐi}) (\textsc{fem} archita \xm{ˈar.ʈ͡ʂi.ta}) ancestor}
\skipline

Markings with \textsc{small caps} are used to present different grammatical
features. \textsc{pl} (plural) is showing you what is the plural form of the
noun, as nouns in Andro are changing with grammatical number.

Marking \textsc{fem} shows you what is the form in the feminine grammatical
gender, as nouns in Andro have assigned grammatical gender. You will also notice
some nouns marked with (\textsc{fem}), which are only existing in feminine
grammatical gender, for example:

\dictwordb{hu͞ekapa}[ˈxuɛ.ka.pa]
\dictterm{n}{(\textsc{fem}) computer}
\skipline

Both markings, for plural and feminine form, are optional.

In case of verbs, you will see the marking \textsc{pst}, showing the past form
of the verb, as verbs conjugate against tenses.

\dictwordb{kuanti}[ˈku.an.ti]
\dictterm{v}{(\textsc{pst} kuant [ˈku.aŋt]) to hunt}
\skipline

In case of adjectives, up to two markings may be presented: \textsc{comp} and
\textsc{supl}, presenting comparative and superlative form, respectively.

\dictwordb{wa͞ime}[ˈwai.mɛ]
\dictterm{adj}{(\textsc{comp} wa͞ime͞a [ˈwai.mɛa], \textsc{supl} wa͞ime͞am [ˈwai.mɛam]) wide}
\skipline

\note{Finally, you may encounter text noted like this. These are usually additional notes about cultural aspects.}
\skipline

\end{spacing}