% \chapter{Appendix B, Examples}
% LTeX: enabled=false

\section{Simple examples}
\glossex{Seja haji.}{seja haji}{Sun shine.PRS}{``The sun shines.''}
\glossex{Seja vi haji.}{seja vi haji}{Sun IPFV shine.PRS}{``The sun is shining.''}
\glossex{Ore Seja haji.}{ore seja haji}{now Sun shine.PRS}{``The sun is shining.''}
\glossex{Seja va hajit.}{seja va haji-t}{sun PFV shine-PST}{``The sun shone.''}
\glossex{Seja ze haji.}{seja ze haji}{sun FUT shine}{``The sun will shine.''}
\glossex{Seja vi hajit.}{seja vi haji-t}{sun IPFV shine-PST}{``The sun has been shining.''}
\glossex{Seja haji rede}{seja haji rede}{sun shine.PRS again}{``The sun is shining again.''}
\glossex{Seja ze haji zetay}{seja ze haji zetay}{sun FUT shine tomorrow}{``The sun will shine tomorrow.''}
\glossex{Seja haji siro.}{seja haji siro}{sun shine.PRS bright.ADV}{``The sun shines brightly.''}
\glossex{Siro Seja haji}{siro seja haji}{bright.ADJ sun shine.PRS}{``The bright sun shines.''}
\glossex{Seja sormi ore.}{seja sormi ore}{sun sunrise.PRS now}{``The sun is rising now.''}
\glossex{Sormi ore.}{sormi ore}{sunrise.PRS now}{``The sun is rising now.''}
\glossex{Recha rujalaros rajit.}{recha rujalar-os raji-t}{all people-PL shout-PST}{``All the people shouted.''}
\glossex{Recha rajit.}{recha raji-t}{all shout-PST}{``All the people shouted.''}
\glossex{Noreja rajit.}{noreja raji-t}{some.people shout-PST}{``Some of the people shouted.''}
\glossex{Vyele rujalaros rajit ka razi.}{vyele rujalar-os raji-t ka razi}{many people-PL shout-PST two times}{``Many of the people shouted twice.''}
\glossex{Osaker rajit ka razi.}{osaker raji-t ka razi}{many.people shout-PST two times}{``Many of the people shouted twice.''}
\glossex{Gusto rujalaros rajiri vicho.}{gusto rujalar-os rajiri vicho}{happy people-PL shout.PRS often}{``Happy people often shout.''}
\glossex{Muchi ka͞ufit ared.}{muchi ka͞ufi-t ared}{kitten jump-PST upward.ADV}{``The kitten jumped up.''}
\glossex{Muchi ka͞ufit on tasek.}{muchi ka͞ufi-t on tasek}{kitten jump-PST on table}{``The kitten jumped onto the table.''}
\glossex{Myi vir muchi uneńit.}{myi vir muchi un-eńi-t}{1SG.POSS little cat away-go-PST}{``My little kitten walked away.''}
\glossex{Aḱame osupi.}{aḱame osupi}{rain fall.PRS}{``It's raining.''}
\glossex{Aḱame osupit.}{aḱame osupi-t}{rain fall-PST}{``The rain came down.''}
\glossex{Muchi labi in aḱame.}{muchi labi in aḱame}{cat play.PRS in rain}{``The kitten is playing in the rain.''}
\glossex{Aḱame aśtot.}{aḱame aśto-t}{rain stop-PST}{``The rain has stopped.''}
\glossex{Aḱame ze aśtopi naári.}{akame ze aśtopi naári}{rain FUT stop soon}{``Soon the rain will stop.''}
\glossex{Avi bochada aḱame ze aśtopi naári.}{avi bochada aḱame ze aśtopi naári}{have.PRS hope rain FUT stop soon}{``I hope the rain stops soon.''}
\glossex{Mosno faji lot volu chet.}{mosno faj-i lo-t volu chet}{wild animal-PL live-PST sometime here}{``Once wild animals lived here.''}
\glossex{Egla rav́at gepo.}{egl-a rav́a-t gepo}{3SG-F look.around-PST slow.ADV}{``Slowly she looked around.''}
\glossex{Ti uneńi do!}{ti uneńi do}{2SG go.away IMP}{``Go away!''}
\glossex{Uneńi do!}{Uneńi do}{go.away IMP}{``Go away!''}
\glossex{Eni heme!}{eni heme}{go CHR}{``Let's go!''}
\glossex{Ti vage eni.}{ti vage eni}{2SG HORT go}{``You should go.''}
\glossex{Vage eni.}{vage eni}{HORT go}{``You should go.''}
\glossex{Ze esi gusto eni na.}{ze esi gusto eni na}{FUT be happy go NMLZ}{``I will be happy going.'' \\ ``I will be happy to go.''}
\glossex{Egli ze iéni naári.}{egli ze i-éni naári}{3SG.M FUT VEN-go soon}{``He will arrive soon.''}
\glossex{Chidoyi moroza unj́aret.}{chido-yi moroza un-j́are-t}{child-POSS ball out-roll-PST}{``The baby's ball has rolled away.''}
\glossex{Ka bovos vi tesuchi.}{ka bov-os vi tesuchi}{two boy-PL IPFV cooperate.PRS}{``The two boys are working together.''}
\glossex{Ka bovos vi fari tesiko.}{ka bov-os vi fari tesiko}{two boy-PL IPFV work.PRS together.ADV}{``The two boys are working together.''}
\glossex{Neb́i͞a ze osupi mireisro.}{neb́i͞a ze osupi mireisro}{mist FUT fall probably}{``This mist will probably clear away.''}
\glossex{Fote͞o flos dolozi in reso.}{fote͞o flo-s dolozi in reso}{lovely.ADJ flower-PL grow.PRS in everywhere}{``Lovely flowers are growing everywhere.''}
\glossex{Noni vage vibi gepe.}{noni vage vibi gep-e}{1PL HORT eat.PRS slow.ADV-COMP}{``We should eat more slowly.''}
\glossex{Ti va iént tar chid.}{ti va i-én-t tar chid}{2SG PFV VEN-go-PST too fast.ADV}{``You have come too soon.''}
\glossex{Ti vipini rizo͞e chiwi.}{ti vipini rizo-e chiwi}{2SG must.PRS.AUX neat-COMP write}{``You must write more neatly.''}
\glossex{Siniko kebuér varsi rizo onvojibo.}{siniko kebuér varsi rizo onvojibo}{wonderful palace stand.PRS exact.ADV opposite.ADV}{``Directly opposite stands a wonderful palace.''}
\glossex{Henryi pelir lukat.}{henr-yi pelir luka-t}{henry-POSS dog lose.way-PST}{``Henry's dog is lost.''}
\glossex{Pelir yi Henri lukat chyi.}{pelir yi henri luka-t chyi}{dog POSS henry lose.way-PST REFL}{``Henry's dog is lost.''}
\glossex{Myi muche esi ruko.}{myi muche esi ruko}{1SG.POSS cat be.PRS black}{``My cat is black.''}
\glossex{Myi muche ruko.}{myi muche ruko}{1SG.POSS cat black}{``My cat is black.''}
\glossex{Loesa yi vir hima esi girefaro.}{loesa yi vir hima esi girefar-o}{doll POSS small girl be.PRS damage-ADJ}{``The little girl's doll is broken.''}
\glossex{Machagi fahojo inuro.}{machagi fahojo inuro}{sleep.PRS loud.ADV usually.ADV}{``I usually sleep soundly.''}
\glossex{Chidos jikit mito Jek.}{chid-os jiki-t mito jek}{child-PL run-PST back.ADV jack}{``The children ran after Jack.''}
\glossex{Epi huhu mibozor labi.}{epi huhu mibozor labi}{can after school play}{``I can play after school.''}
\glossex{Noni ent o kavala per tamoki.}{noni en-t o kavala per tamoki}{1PL go-PST to village for visit.PRS}{``We went to the village for a visit.''}
\glossex{Noni iént o aḱumar.}{noni i-én-t o aḱumar}{1PL VEN-go-PST to river}{``We arrived at the river.''}
\glossex{Va ingek a ti}{va ingek a ti}{PFV wait.PST on 2SG}{``I have been waiting for you.''}
\glossex{Naberos set ra͞o nafrita.}{naber-os set rao na-frita}{camper-PL sit.PST around camp-fire}{``The campers sat around the fire.''}
\glossex{Vir hima e il muchi set ne ne͞a mi.}{vir hima e il muchi set nea mi}{small girl and 3SG.POSS kitten sit.PST near.physically 1SG}{``A little girl with a kitten sat near me.''}
\glossex{Chido ingek a il vapal ner ostro.}{chido ingek a il vapal ner ostro}{child wait.PST on 3SG.POSS father near.physically door}{``The child waited at the door for her father.''}
\glossex{Kavala yi vyekka hima mant il muchi asweo tay.}{kavala yi vyekka hima man-t il muchi asweo tay}{village POSS old.SUPL girl lose-PST 3SG.POSS kitten previous day}{``Yesterday the oldest girl in the village lost her kitten.''}
\glossex{Kavala yi vyekka hima yi muchi lukat asweo tay.}{kavala yi vyekka hima yi muchi lukat asweo tay}{village POSS old.SUPL girl POSS kitten lose.way-PST previous day}{``Yesterday the oldest girl in the village lost her kitten.''}
\glossex{In che kavala ti uńatal?}{in che kavala ti uńatal}{in DEM village 2SG be.born-PST}{``Were you born in this village?''}
\glossex{Lipe tyi brat leehi?}{lipe tyi brat leehi}{good.ADV 2SG.POSS brother dance}{``Can your brother dance well?''}
\glossex{Sinit rujaler?}{sini-t rujaler}{leave-PST man}{``Did the man leave?''}
\glossex{A ti tyi sora iéni?}{a ti tyi sora iéni}{for 2SG 2SG.POSS sister come}{``Is your sister coming for you?''}
\glossex{Zetay ti epi iéni?}{zetay ti epi iéni}{tomorrow 2SG can.AUX come}{``Can you come tomorrow?''}
\glossex{Oban mikyimarida ne͞areros va sićherit?}{oban mikyimarida ne͞arer-os va sićheri-t}{for.the.time winter neighbor-PL PFV leave-PST}{``Have the neighbors gone away for the winter?''}
\glossex{Feban aḱame kahuna kanti?}{feban aḱame kahuna kanti}{during rain bird sing.PRS}{``Does the robin sing in the rain?''}
\glossex{O kijursa a noni ti ze egi?}{o kijursa a noni ti ze egi}{to concert with 1PL 2SG FUT go}{``Are you going with us to the concert?''}
\glossex{Volumo churche dalaka ti vayarenit?}{volumo churche dalaka ti vayareni-t}{anywhen.Q through jungle 2SG travel-PST}{``Have you ever travelled in the jungle?''}
\glossex{Noni fenit aḱumaryi mered churche vyele kilometerdi.}{noni feni-t aḱumar-yi mered churche vyele kilometer-di}{1PL swim-PST river-POSS down through many kilometer-PL}{``We sailed down the river for several miles.''}
\glossex{Recha saperi on kuanti.}{recha saperi on kuanti}{everybody know.PRS about hunt.PRS}{``Everybody knows about hunting.''}
\glossex{In sejo zevoyim huhu tayav́al noni sint́epit in abarji.}{in sejo zevoyim huhu tayav́al noni si-nt́epi-t in abar-ji}{in sunny morning after solstice 1PL off-go-PST in mountain-PL}{``On a Sunny morning after the solstice we started for the mountains.''}
\glossex{Tom chikat dyet chichamayi virfarji.}{tom chika-t dyet chichama-yi virfar-ji}{tom laugh-PST because.of monkey-POSS trick-PL}{``Tom laughed at the monkey's tricks.''}
\glossex{Vek rujalar one enyasula vajat dyet duloge.}{vek rujalar one enyasula vajat dyet duloge}{old man with walking.stick stand.PST behind fence}{``An old man with a walking stick stood beside the fence.''}
\glossex{Sukirrayi sehola est kopero da sif́emo femji.}{sukirra-yi sehola est kopero da sif́em-o fem-ji}{squirrel-POSS nest be.PST hidden INS hang-ADJ branch-PL}{``The squirrel's nest was hidden by drooping boughs.''}
\glossex{Vir granos ingek tranoko un miker a tepa novekayi seja.}{vir gran-os inge-k tranoko un miker a tepa noveka-yi seja}{tiny seed-PL wait-PST patiently.ADV under snow for warm spring-POSS sun}{``The little seeds waited patiently under the snow for the warm spring sun.''}
\glossex{Vyele vir hinji one floyi topeka a egyi buutji va leehit ra͞o nafrita.}{vyele vir hin-ji one flo-yi topeka-s a egyi buut-ji va leehi-t ra͞o nafrita}{many little little.girl-PL with flower-POSS wreath-PL physically.on 3PL.POSS head-PL PFV dance-PST around bonfire}{``Many little girls with wreaths of flowers on their heads danced around the bonfire.''}
\glossex{Iḱopera yi rakel osupet a busama.}{iḱopera yi rakel osup-et a busama}{cover POSS basket fall-PST on floor}{``The cover of the basket fell to the floor.''}
\glossex{Yelonayi ati bove aśtot ne͞a meńi.}{yelona-yi a-ti bove aśto-t ne͞a meńi}{queue-POSS one-ORD buy stop-PST near entrance}{``The first boy in the line stopped at the entrance.''}
\glossex{A veray yi varda vek sago hinaja lot in vir wamaga.}{a veray yi varda vek sago hinaja lot-ø in vir wamaga}{on top POSS hill old wise woman live-PST in old hut}{``On the top of the hill in a little hut lived a wise old woman.''}
\glossex{Feban niyi ewagi na a kavala enidi a filetos vicho.}{feban niyi ewagi na a kavala enidi-t a filet-os vicho}{during 1PL.POSS stay NMLZ on village walk-PST on field-PL often.ADV}{``During our residence in the country we often walked in the pastures.''}
\glossex{Voli tyi koaleros a͞u espero ze iéni?}{voli tyi koaler-os a͞u espero ze iéni}{when.Q 2SG.POSS guest-PL from city FUT arrive}{``When will your guests from the city arrive?''}
\glossex{Ner aḱumaryi sina, chyi larima vidi taryo a sormo.}{ner aḱumar-yi sina, chyi larima vidi taryo a sormo}{near.physically river-POSS exit 3.INAN.POSS road turn sharp.ADV on east}{``Near the mouth of the river, its course turns sharply towards the East.''}
\glossex{Ka ardo abarji bugi menate mereda mede.}{ka ardo abar-ji bugi menate mereda mede}{two high mountain-PL lay fertile valley between.ADV}{``Between the two lofty mountains lay a fertile valley.''}
\glossex{}{}{}{``Among the wheat grew tall red poppies.''}
\glossex{}{}{}{``The strong roots of the oak trees were torn from the ground.''}
\glossex{}{}{}{``The sun looked down through the branches upon the children at play.''}
\glossex{}{}{}{``The west wind blew across my face like a friendly caress.''}
\glossex{}{}{}{``The spool of thread rolled across the floor.''}
\glossex{}{}{}{``A box of growing plants stood in the Window.''}
\glossex{Mi zosu gusto.}{mi zosu gusto}{1SG very happy}{``I am very happy.''}
\glossex{}{}{}{``These oranges are juicy.''}
\glossex{}{}{}{``Sea water is salty.''}
\glossex{}{}{}{``The streets are full of people.''}
\glossex{}{}{}{``Sugar tastes sweet.''}
\glossex{Seiti ifrit votepo.}{seiti ifrit votepo}{feel.PRS fire hot}{``The fire feels hot.''}
\glossex{Mi seiti ifrit esi votepo.}{mi seiti ifrit esi votepo}{1SG feel.PRS fire be.PRS hot}{``The fire feels hot.''}
\glossex{Vir hina epit ziralo rori.}{vir hina epi-t ziralo rori}{little girl might-PST lonely think.AUX}{``The little girl seemed lonely.''}
\glossex{Vir boveyi vapal est suyer volu.}{vir bove-yi vapal es-t suyer volu}{little boy-POSS father be-PST sailor somewhen}{``The little boy's father had once been a sailor.''}
\glossex{}{}{}{``I have lost my blanket.''}
\glossex{}{}{}{``A robin has built his nest in the apple tree.''}
\glossex{}{}{}{``At noon we ate our lunch by the roadside.''}
\glossex{Jones-epié simoset saler for egi vir bove.}{jones-epié simose-t saler for il vir bove}{jones-HON create-PST knife DAT 3SG.POSS small boy}{``Mr. Jones made a knife for his little boy.''}
\glossex{Egyi inerji epi rori zosu gusto.}{egyi iner-ji epi.rori zosu gusto}{3PL.POSS voice-PL seem.to.be very happy}{``Their voices sound very happy.''}
\glossex{Egyi inerji inerai zosu gusto.}{egyi iner-ji inerai zosu gusto}{3PL.POSS voice-PL sound very happy}{``Their voices sound very happy.''}
\glossex{Atay hetay esi?}{a-tay hetay esi}{first-day today be.PRS}{``Is today Monday?''}
\glossex{A͞u laák recha ayeri va osupit?}{au laák recha ayer-i va osupi-t}{from tree all leave-PL PFV fall.down-PST}{``Have all the leaves fallen from the tree?''}
\glossex{Medo voli baán ze esi?}{medo voli baán ze esi}{ready when time FUT be}{``Will you be ready on time?''}
\glossex{Missani fo mi ti ze siyurchi?}{missani fo mi ti ze siyurchi}{message DAT 1SG 2SG FUT send}{``Will you send this message for me?''}
\glossex{Vi ingeki a mi?}{vi ingeki a mi}{IPFV wait for 1SG}{``Are you waiting for me?''}
\glossex{}{}{}{``Is this the first kitten of the litter?''}
\glossex{Tur gruwe fo ti je kutos esi?}{tur gruwe fo yi je kutos esi}{too big DAT 2SG DEM shoe.PL be.PRS}{``Are these shoes too big for you?''}
\glossex{Ali wa͞ime aḱumar esi?}{ali wa͞ime aḱumar esi}{how.much.Q wide river be.PRS}{``How wide is the River?''}
\glossex{Vage kaéti.}{vage kaéti}{HORT listen}{``Listen.''}
\glossex{Kaéti do.}{kaéti do}{listen IMP}{``Listen.''}
\glossex{Seysi chet ne͞a myi do.}{seysi chet ne͞a myi do}{sit here near 1SG.ACC IMP}{``Sit here by me.''}
\glossex{}{}{}{``Keep this secret until tomorrow.''}
\glossex{Tesiko noni eni heme.}{tesiko noni eni heme}{together 1PL go CHR}{``Come with us.''}
\glossex{}{}{}{``Bring your friends with you.''}
\glossex{}{}{}{``Be careful.''}
\glossex{}{}{}{``Have some tea.''}
\glossex{Pip e il pelir est lipe alyes.}{Pip e il pelir est lipe alye-s}{Pip and 3SG.POSS dog be.PST good friend-PL}{``Pip and his dog were great friends.''}
\glossex{Jon e Elizabet esi brat e sora}{Jon e Elizabet esi brat e sora}{John and Elizabeth be.PRS brother and sister}{``John and Elizabeth are brother and sister.''}
\glossex{Ti e mi ze eni tesiko.}{ti e mi ze eni tesiko}{2SG and 1SG FUT go together.ADV}{``You and I will go together.''}
\glossex{Ego͞i id́ak recha ostros e dakos.}{ego͞i id́ak recha ostro-s e dako-s}{3PL open.PST every door-PL and window-PL}{``They opened all the doors and windows.''}
\glossex{Egi esi vir, abe valor.}{egi esi vir abe valor}{3SG be small but strong}{``He is small, but strong.''}
\glossex{}{}{}{``Is this tree an oak or a maple?''}
\glossex{Arso zor stobo ari esi?}{arso zor stobo ari esi}{azure exclusive.or gray sky be.PRS}{``Does the sky look blue or gray?''}
\glossex{Vage iéni a vapal yen natali͞a.}{vage iéni a vapal yen natali͞a}{HORT come together.with father or mother}{``Come with your father or mother.''}
\glossex{Mi ge payto esi, abe zosu gusto.}{mi ge payt-o esi abe zosu gusto}{1SG PASS tire-ADJ but very happy}{``I am tired, but very happy.''}
\glossex{}{}{}{``He played a tune on his wonderful flute.''}
\glossex{}{}{}{``Toward the end of August the days grow much shorter.''}
\glossex{}{}{}{``A company of soldiers marched over the hill and across the meadow.''}
\glossex{Ati taúnin yi yeitto zosu intrise.}{a-ti taúnin yi yeitto zosu intrise}{one-ORD part POSS story very interesting}{``The first part of the story is very interesting.''}
\glossex{}{}{}{``The crow dropped some pebbles into the pitcher and raised the water to the brim.''}
\glossex{}{}{}{``The baby clapped her hands and laughed in glee.''}
\glossex{Fiji aśtopi e esi yubo do.}{fiji astopi e esi yubo do}{game.PRS stop.PRS and be.PRS quiet.ADV IMP}{``Stop your game and be quiet.''}
\glossex{}{}{}{``The sound of the drums grew louder and louder.''}
\glossex{Samos zor mikyimarida ti anvi?}{samos ji mikyimarida ti anvi}{summer or winter 2SG prefer}{``Do you like summer or winter better?''}
\glossex{}{}{}{``That boy will have a wonderful trip.''}
\glossex{}{}{}{``They popped corn, and then sat around the fire and ate it.''}
\glossex{}{}{}{``They won the first two games, but lost the last one.''}
\glossex{}{}{}{``Take this note, carry it to your mother; and wait for an answer.''}
\glossex{}{}{}{``I awoke early, dressed hastily, and went down to breakfast.''}
\glossex{Aha! Va kanit ti!}{aha va kani-t ti}{aha PFV catch-PST 2SG}{``Aha! I have caught you!''}
\glossex{}{}{}{``This string is too short!''}
\glossex{}{}{}{``Oh, dear! the wind has blown my hat away!''}
\glossex{}{}{}{``Alas! that news is sad indeed!''}
\glossex{}{}{}{``Whew! that cold wind freezes my nose!''}
\glossex{}{}{}{``Are you warm enough now?''}
\glossex{}{}{}{``They heard the warning too late.''}
\glossex{}{}{}{``We are a brave people, and love our country.''}
\glossex{Recha chidos aygo Meri iént.}{recha chid-os aygo Meri iént}{all.of child-PL apart.from Mary come.PST}{``All the children came except Mary.''}
\glossex{}{}{}{``Jack seized a handful of pebbles and threw them into the lake.''}
\glossex{}{}{}{``This cottage stood on a low hill, at some distance from the village.''}
\glossex{}{}{}{``On a fine summer evening, the two old people were sitting outside the door of their cottage.''}
\glossex{}{}{}{``Our bird's name is Jacko.''}
\glossex{}{}{}{``The river knows the way to the sea.''}
\glossex{}{}{}{``The boat sails away, like a bird on the wing.''}
\glossex{}{}{}{``They looked cautiously about, but saw nothing.''}
\glossex{}{}{}{``The little house had three rooms, a sitting room, a bedroom, and a tiny kitchen.''}
\glossex{}{}{}{``We visited my uncle's village, the largest village in the world.''}
\glossex{}{}{}{``We learn something new each day.''}
\glossex{}{}{}{``The market begins five minutes earlier this week.''}
\glossex{}{}{}{``Did you find the distance too great?''}
\glossex{}{}{}{``Hurry, children.''}
\glossex{}{}{}{``Madam, I will obey your command.''}
\glossex{}{}{}{``Here under this tree they gave their guests a splendid feast.''}
\glossex{}{}{}{``In winter I get up at night, and dress by yellow candlelight.''}
\glossex{}{}{}{``Tell the last part of that story again.''}
\glossex{}{}{}{``Be quick or you will be too late.''}
\glossex{}{}{}{``Will you go with us or wait here?''}
\glossex{}{}{}{``She was always, shabby, often ragged, and on cold days very uncomfortable.''}
\glossex{}{}{}{``Think first and then act.''}
\glossex{}{}{}{``I stood, a little mite of a girl, upon a chair by the window, and watched the falling snowflakes.''}
\glossex{}{}{}{``Show the guests these shells, my son, and tell them their strange history.''}
\glossex{}{}{}{``Be satisfied with nothing but your best.''}
\glossex{}{}{}{``We consider them our faithful friends.''}
\glossex{}{}{}{``We will make this place our home.''}
\glossex{}{}{}{``The squirrels make their nests warm and snug with soft moss and leaves.''}
\glossex{}{}{}{``The little girl made the doll's dress herself.''}
\glossex{}{}{}{``I hurt myself.''}
\glossex{}{}{}{``She was talking to herself.''}
\glossex{}{}{}{``He proved himself trustworthy.''}
\glossex{}{}{}{``We could see ourselves in the water.''}
\glossex{}{}{}{``Do it yourself.''}
\glossex{}{}{}{``I feel ashamed of myself.''}
\glossex{}{}{}{``Sit here by yourself.''}
\glossex{}{}{}{``The dress of the little princess was embroidered with roses, the national flower of the Country.''}
\glossex{}{}{}{``They wore red caps, the symbol of liberty.''}
\glossex{}{}{}{``With him as our protector, we fear no danger.''}
\glossex{}{}{}{``All her finery, lace, ribbons, and feathers, was packed away in a trunk.''}
\glossex{}{}{}{``Light he thought her, like a feather.''}
\glossex{}{}{}{``Every spring and fall our cousins pay us a long visit.''}
\glossex{}{}{}{``In our climate the grass remains green all winter.''}
\glossex{}{}{}{``The boy who brought the book has gone.''}
\glossex{}{}{}{``These are the flowers that you ordered.''}
\glossex{}{}{}{``I have lost the book that you gave me.''}
\glossex{}{}{}{``The fisherman who owned the boat now demanded payment.''}
\glossex{}{}{}{``Come when you are called.''}
\glossex{}{}{}{``I shall stay at home if it rains.''}
\glossex{}{}{}{``When he saw me, he stopped.''}
\glossex{}{}{}{``Do not laugh at me because I seem so absent minded.''}
\glossex{}{}{}{``I shall lend you the books that you need.''}
\glossex{}{}{}{``Come early next Monday if you can.''}
\glossex{}{}{}{``If you come early, wait in the hall.''}
\glossex{}{}{}{``I had a younger brother whose name was Antonio.''}
\glossex{}{}{}{``Gnomes are little men who live under the ground.''}
\glossex{}{}{}{``He is loved by everybody, because he has a gentle disposition.''}
\glossex{}{}{}{``Hold the horse while I run and get my cap.''}
\glossex{}{}{}{``I have found the ring I lost.''}
\glossex{}{}{}{``Play and I will sing.''}
\glossex{}{}{}{``That is the funniest story I ever heard.''}
\glossex{}{}{}{``She is taller than her brother.''}
\glossex{}{}{}{``They are no wiser than we.''}
\glossex{}{}{}{``Light travels faster than sound.''}
\glossex{}{}{}{``We have more time than they.''}
\glossex{}{}{}{``She has more friends than enemies.''}
\glossex{}{}{}{``He was very poor, and with his wife and five children lived in a little low cabin of logs and stones.''}
\glossex{}{}{}{``When the wind blew, the traveler wrapped his mantle more closely around him.''}
\glossex{}{}{}{``I am sure that we can go.''}
\glossex{}{}{}{``We went back to the place where we saw the roses.''}
\glossex{}{}{}{``"This tree is fifty feet high," said the gardener.''}
\glossex{}{}{}{``I think that this train leaves five minutes earlier today.''}
\glossex{}{}{}{``My opinion is that the governor will grant him a pardon.''}
\glossex{}{}{}{``Why he has left the city is a mystery.''}
\glossex{}{}{}{``The house stands where three roads meet.''}
\glossex{}{}{}{``He has far more money than brains.''}
\glossex{}{}{}{``Evidently that gate is never opened, for the long grass and the great hemlocks grow close against it.''}
\glossex{}{}{}{``I met a little cottage girl; she was eight years old, she said.''}
\glossex{Noni ge siyurchet per karli chu e͞iger e zaálta.}{noni ge siyurche-t per karli chu e͞iger e zaálta}{1PL PASS send-PST for die.PRS ACC emperor and nation}{``We were sent to die for the emperor and nation.''}
\glossex{Abejar voli koboviros karli, egyi natalji inrit.}{abejar voli kobovir-os kar-et, egyi natalji inri-t}{however when soldier-PL die-PST 3PL.POSS mother-PL call-PST}{``But when the soldiers were dying, called out to their mothers.''}
\glossex{Moĺi kaét sotak chu egi e͞iger e zaálta vi inrit.}{moĺi kaét sotak chu egi e͞iger e zaálta vi inri-t}{never hear.PST someone ACC 3SG emperor and nation IPFV call-PST}{``I never heard anyone calling the emperor and the nation.''}
\glossex{Vage loti lipe lot.}{vage loti lipe lot}{HORT live.PRS good life}{``Live a good life.''}
\glossex{Miam ora͞i feśidari e licho esi vimi ze siti no ali ti horo est.}{miam ora͞i feśidari e licho esi vimi ze siti no ali ti horo est}{if god.PL exist.PRS and just be.PRS then FUT interest NEG how.many 2SG devout be.PST}{``If there are gods and they are just, then they will not care how devout you have been.''}
\glossex{Miam ora͞i feśidari, abe nolicho, vimi ti egyi ori vage teviti no.}{miam ora͞i feśidari abe no-licho vimi ti egyi ori vage teviti no}{if god.PL exist but NEG-just then 2SG 3PL.ACC worship HORT want NEG}{``If there are gods, but unjust, then you should not want to worship them.''}
\glossex{Miam ora͞i feśidari no, vimi ti uneńit, abejar vi lot arujaro lot chu ze vi feśidari in saypes yi tyi koólerji.}{miam ora͞i feśidari no vimi ti ze uneńi abejar vi lot arujaro lot chu ze vi feśidari in saype-s yi tyi koóler-ji}{if god.PL exist NEG then 2SG FUT leave however IPFV live.PST noble.ADV ACC IPFV exist in memory-PL POSS 2SG.POSS loved.one-PL}{``If there are no gods, then you will be gone, but will have lived a noble life that will live on in the memories of your loved ones.''}
\glossex{Baljejet joifina flos no.}{baljeje-t joifina flo-s no}{receive-PST kick-NMLZ flower-PL instead.of}{I got a kick instead of flowers.}
\glossex{Akaro arios sormi an kavala ge dozit da ichera.}{akar-o arios sormi an kavala ge dozi-t da ichera}{red-ADJ Moon rise.PRS over village PASS make.dirty-PST INS blood}{``The red Moon is rising over a village smeared with blood.''}
\glossex{Tyi gont ze stani no o voli tay in chu recha vireji uf́riti ze iéni.}{tyi gont ze stani no o=voli tay in chu recha vire-ji ze uf́riti ze i-éni}{2SG.POSS soul FUT rest.PRS NEG until day LOC GEN all star-PL FUT burn.out FUT VEN-come}{``Your soul will not rest until comes the day in which all stars will burn out.''}
\glossex{Saḱaywa yaro lot ti vivi per ti ruókeyi yarji eédi?}{sa-ḱa-y-wa yar-o lot ti vivi per ti ruóke-yi yar-ji eédi}{3-2-times-0 year-ADJ life 2SG to.intend in.order.to 2SG crow-POSS year-PL get.to.know}{``Are you going to live three hundred years to learn the age of the crow?''}
\glossex{Egla esi merlio fo mi.}{egl-a esi merlio fo mi}{3SG.F be.PRS appealing.ADJ DAT 1SG}{``She is pretty in my opinion.'' \\ ``She is appealing to me'' \\ ``I like her.''}
\glossex{Egla merlio fo mi.}{egl-a merlio fo mi}{3SG.F appealing.ADJ DAT 1SG}{``She is pretty in my opinion.'' \\ ``She is appealing to me'' \\ ``I like her.''}
\glossex{Egla merlio mi.}{egl-a merlio mi}{3SG.F appealing.ADJ 1SG}{``She is pretty in my opinion.'' \\ ``She is appealing to me'' \\ ``I like her.''}
\glossex{Egla merlio.}{egl-a merlio}{3SG.F appealing.ADJ}{``She is pretty in my opinion.'' \\ ``She is appealing to me'' \\ ``I like her.''}
\glossex{Mi ya egla merlio.}{mi ya egl-a merlio}{1SG TOP 3SG.F appealing.ADJ}{``As for my opinion, she is pretty.'' \\ ``She is pretty in my opinion.''}

\section{Longer texts}
\subsection{Universal Declaration of Human Rights, Article 1}
\glossex{Recha rujalaros uńatali jarano e chant chane͞o egyi altomi e egyi peremaji.}{recha rujalar-os unatali jarano e chant chaneo egyi altom-i e egyi liche-s}{every man-PL is.born.PRS free.ADJ and.CONJ equal in.terms.of 3PL.POSS honor-PL and.CONJ 3PL.POSS human.right-PL}{``All men are born free and equal in terms of their own dignity and human rights.''}
\glossex{Ego͞i elazi kape e fataba, ego͞i vage dusteri chu bajekeros koy rujalaka.}{ego͞i elazi kape e fataba, vage dusteri chu bajeker-os koy rujalaka}{3PL possess mind and.CONJ conscience EXH act GEN other.people-PL in.a.spirit.of community}{``They are in posession of mind and conscience and should act towards other people in a spirit of community.''}